A Survey of Physical Design Techniques of Information Systems

### 2010-07-25 trunk/cs848-db/Paper.tex
integrated with Sarah's latest work

\documentclass[12pt,a4paper]{article}
\usepackage{fullpage}
\usepackage{multirow}

\usepackage{graphics}
\usepackage{color}
\usepackage[pdftex]{hyperref}
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{url}

\begin{document}

@@

Relational Database Management Systems (RDBMS) have been the most popular type of Information Systems for decades. When we think of designing a database
system, we usually think of which data we need to keep track of and how will we properly organize it into meaningful tables. However, this is not the only phase
in database design. All of the designed tables need to be physically stored on disk. How these tables will be stored and accessed greatly affects the database
performance. Physical database design is concerned with the actual storage of data on disk. This means how the files are organized, and how they are accessed.
Data is stored in the
form of records in files which reside on a physical disk. In this sense, we try to find the most efficient way possible to store and access the data such that
queries are answered and executed efficiently. In the traditional relational database system, this involved selecting proper indexing techniques,
partitioning the data for more efficient access, materializing the results of some queries, clustering and compressing the data, and many other
techniques~\cite{fikelstein1988physical,lightstone2007physical}. The physical design aspects of relational databases have been studied in much detail. However,
over time, different types of Information Systems have emerged that have different characteristics. For example, most relational databases have been disk-based
where the primary copy of the data resides on disk. This is as opposed to Main Memory Relational Databases (MMDB) where the primary copy of the data resides in
main memory and no disk operations are needed to query the data. Such differences pose new requirements for the physical design of these systems.

@@

The rest of the survey is organized as follows. Section~\ref{SEC-ELEMENTS} first explains different elements of physical database design used in traditional
disk-based databases. Section~\ref{SEC-DIFFSYS} then discusses how each of the three other systems uses these elements, as
applicable. Sections~\ref{SEC-MMDB},~\ref{SEC-XML},~\ref{SEC-WAREHOUSES}, discuss Main Memory Databases, XML Databases, and Data Warehouses
respectively. Throughout these sections, comparisons are made as to how the different elements have been modified to fit the needs of each system.
Section~\ref{SEC-OPEN} mentions some of the
open problems of physical database design, and suggests possible solutions. Finally Section~\ref{SEC-CONCL} summarizes and concludes this survey.

@@

Relational Databases were first introduced by Codd~\cite{codd1970relational}. Disk based Relational Database Management Systems (RDBMS) are the most common and
popular
type of Information Systems. Since they have been extensively researched and have been around for many decades, we use them as the basis of our comparison. In
this section, we describe the different elements of physical design in RDBMS. These elements will also be examined in the other systems to see how they apply
to them.

@@

Btrees (short for Balanced Trees) are the most used database index structure.  The Btree is essentially similar to the traditional Binary Search Tree, but
instead of having one value per node, a B-tee can have many values per node~\cite{comer1979ubiquitous}. B-trees are suitable for relational databases since the
cost of retrieval in a B-tree is at most proportional to: $log_{d}\frac{n+1}{2}$. The cost of insertion
and deletion is at most proportional to $log_{d}n$ due to the possibility of progressing back up the tree to balance it after an insertion or a deletion.
Although B-trees do well in retrieval, deletion and insertion, they do not perform well in sequential search.

Different variations of B-trees have been used. For example, B+trees are now the main method of indexing in current disk-based Relational
Databases~\cite{lightstone2007physical} such as DB2, Oracle, and SQL Server. The main difference between the
B+tree and the Btree is that only leaf nodes contain data pointers in a B+tree, and leaf nodes contain pointers to each other as shown in
Figure~\ref{fig:b+tree}. In Prefix B+trees, instead of using the
actual key value of a record, a prefix from the key value is used to decrease the storage size needed which will in turn allow more records to be stored per
node thus decreasing the height of the tree. B+trees have been successful in relational databases since data is ultimately recorded in files. Since B+trees have
a high fanout, they allow less I/O operations to access a specific data record which is stored in a specific file.

Of course, there are other indexing techniques used such as hash tables, and bitmap indices. EXPLAAAAAAAAIN

briefly discuss clustered vs. unclustered ?

@@

In a large database system, there are usually a few complex queries that require the joining of many large tables. If these queries are frequently run, then the
database performance is likely to suffer, and users have to wait for a long time to get the results back. If we know these queries beforehand, it makes sense to
simply store the results of the queries on disk instead of recomputing them each time. Oracle 8i introduced materialized views to fit this purpose. By
precalculating the
results of a complex query, and storing them in a table on disk, the new table with the results is very likely to be much smaller than the original tables
decreasing I/O access costs and thus increasing performance~\cite{lightstone2007physical, chaudhuri1998overview}. Once a materialized view exists, the user can
either explicitly query the materialized view, or enter the original query, and have the query optimizer rewrite the query to use the materialized views instead
of the original tables. Many query optimization techniques have relied on rewriting queries using views (E.g.~\cite{levy1995answering, gupta1995aggregate,
goldstein2001optimizing,
abiteboul1998complexity}), but discussing the details of the rewriting process is beyond the scope of this survey.

@@

Partitioning is an important aspect of physical design as it reduces table scanning time~\cite{lightstone2007physical}. There are two main categories of
partitioning: horizontal partitioning
and vertical partitioning. Horizontal partitioning divides the table into sets of table rows where each row or record still has all the attributes of the table.
For example, dividing the data by date where all data dating less than year 2000 lies in one partition while all data dating more lies in another. On the other
hand, vertical partitioning reduces the width of the dataset by storing some attributes in one table, and some in another. Usually, less frequently queried
columns are partitioned together to reduce the size of the table.

The main advantage of any type of partitioning is to reduce the amount of time it takes to scan a table which in turn improves performance. For example, if a
table has 1000 records, and a query is only using the first 20 records then having these 20 records in a partition containing 50 records will save time as
opposed to examining all the 1000 records. Horizontal partitioning is more widely used, and is implemented in all current DBMSes. Table~\ref{tab:horpartition}
shows different categories of horizontal partitioning, and the situations they are suited to. Combining partitioning and indexing can often lead to better
performance. For example, if List partitioning is used to divide a table by regions and then a bitmap index is used on that table, finding records in the same
region or state can be much faster. Finally, partitioning the data into different physical devices to achieve parallelism when accessing it can greatly
increase the performance of the database.

NEED SOME REFERENCES HERE

\begin{table}[!t]
\scalebox{0.8}{
 \begin{tabular}{p{2cm}||p{4cm}|p{4cm}|p{4cm}|p{4cm}}
& \textbf{Range Partitioning} & \textbf{List Partitioning} & \textbf{Hash Partitioning} & \textbf{Composite Partitioning}\\
\hline\hline
\textbf{How?} & If key falls in range & List of keys & Hash function& Hybrid\\
\hline
\textbf{Why?} & Disk limits & Group point values & Range/List are N/A & According to needs\\
\hline
\textbf{Benefits} & \multicolumn{4}{c}{Partition elimination, Improved administration, Fast roll-in/out, and Implicit clustering }
\\
\hline
\textbf{Example} & \begin{itemize}
                    \item Zip code 
		    \item Time range
                   \end{itemize} &
		    \begin{itemize}
		    \item Regions
		    \item States
		    \end{itemize} &
		    \begin{itemize}
		    \item EmployeeId
		    \item ProductId
		    \end{itemize} & Hash then List \\
\hline
\textbf{RDBMS} & MySQL, DB2, Oracle, PostgreSQL, SQL Server & MySQL, Oracle, PostgreSQL, SQL Server & MySQL, Oracle & MySQL, Oracle\\

\end{tabular}
}
\caption{Types of horizontal partitioning}
\label{tab:horpartition}
\end{table}

@@

A Main Memory Database (MMDB) (sometimes referred to as in-memory database) is one where the primary copy of the data resides in the main memory of the system
rather than on a
disk~\cite{garcia1992main}. It should be noted that this is different from caching. Disk based database systems use the main memory to cache query results.
However, MMDB's primary copy of the data resides in main memory. Accessing data in main memory is faster than disk, and so MMDBs usually have better
performance that disk based systems. However, main memory is more expensive and the size is limited in comparison to disk. However, MMDBs are essential to use
where some of the collected data cannot, or will not, be stored on disk in the first place. For example, phone switch data may be needed for reasoning with
other stored data, but it is captured in real time and will not necessarily be stored on disk. MMDB have many implementation challenges that have been addressed
throughout the research community. These challenges will be discussed throughout this section. Table~\ref{tab:mmdbsumm} summarizes
the physical design elements of MMDBs.

\begin{table}[t!]
\centering
\scalebox{0.7}{
\begin{tabular}{p{3cm}|p{6cm}|p{6cm}}
\hline
 \textbf{Physical Design Element} & \textbf{Desired Features} & \textbf{Methods/Structures Used}\\
\hline\hline
Index Structures & \begin{itemize}
                    \item No need to store actual values in index \item Larger node size that is aligned with cache line size for cache consciousness
                   \end{itemize} & \begin{itemize}
                                    \item B Tree
\item B+ Tree
\item CSB+ Tree
\item pB+ Tree
\item T Tree
\item CST Tree
\item CSS Tree
                                   \end{itemize}\\
\hline
Materialized Views & \begin{itemize} \item Not highly utilized since processing and memory access is cheap \end{itemize}& N/A\\
\hline
Partitioning & \begin{itemize} \item Only needed in secondary storage to speed up reloading in case of a crash \end{itemize} & \begin{itemize}
                                                                                            \item Horizontal Partitioning
\item Vertical Partitioning
                                                                                           \end{itemize}\\
\hline
Clustering & \begin{itemize} \item Not needed since random or dispersed main memory access is not more expensive than sequential access \end{itemize}
& N/A

@@

In this section, we talk about three families of index structures used in MMDB: B Trees, T Trees, and Binary Search Trees. There are more recent index
structures developed for MMDB such as J+ Trees~\cite{luan2009prefetching}, but which have not gained much popularity, and so we do not discuss them in our
survey. Table~\ref{tab:mmdbindexsumm} summarizes the features of the different index structures discussed in this section with respect to MMDB. The T Tree is
the most used index for MMDB (E.g. MySQL CLuster). However, there are other indexes used as well. For example, IBM's solidDB uses Trie (or prefix tree) for its
indexing. For their purposes, it served as a good index for main memory since it eliminates the need for many key comparisons~\cite{ibmsoliddb}.

\begin{table}[t!]
\centering
\scalebox{0.7}{
\begin{tabular}{p{2cm}|p{6cm}|p{6cm}}
\hline
 \textbf{Index Structure} & \textbf{Features} & \textbf{Cache Consciousness}\\
\hline\hline
B Trees &
\begin{itemize}\item Good storage utilization
\item Reasonably quick searching
\item Fast updating
\end{itemize} 
& \begin{itemize}\item Reasonable cache behavior if node fits in cache line\end{itemize}\\
\hline
B+ Trees & 
\begin{itemize}
\item Reasonably quick searching
\item Fast updating
\end{itemize} &
\begin{itemize}
\item Reasonable cache behavior if node fits in cache line\end{itemize}\\
\hline
CSB+ Trees & \begin{itemize}
              \item Same features of B+ Trees
\item Stores child nodes in an array. Stores pointer to first child only and uses index calculation to get the rest
             \end{itemize} &
\begin{itemize}
 \item Good cache performance due to improved locality \& lower tree height (because of larger node size)
\end{itemize}\\
\hline

@@

\subsection{Data Warehouses}
\label{SEC-WAREHOUSES}
A data warehouse is a subject-oriented, integrated, time-variant, and non-volatile collection of data and decision support technologies, aimed at enabling the knowledge worker (executive, manager, analyst) to make better and faster decisions \cite{560407, 248616}. More than half of IT executives named Data warehousing as the highest-priority post-millennium project for them \cite{sen2005comparison}. The value of data warehousing for an organization depends on the organization's need for reliable, consolidated, unique and integrated reporting and analysis of its data, at different levels of aggregation.

Data warehousing has been shown useful in many industries: manufacturing (for order shipment and customer support), retail (for user profiling and inventory
management), financial services (for claims analysis, risk analysis, credit card analysis, and fraud detection), transportation (for fleet management),
telecommunications (for call analysis and fraud detection), utilities (for power usage analysis), and healthcare (for outcomes analysis) \cite{248616}.

@@

Although data in a data warehouse is extracted and loaded from multiple on-line transaction processing (OLTP) data sources (including DB2, Oracle, IMS databases, and flat files) using Extract, Transfer, and Load (ETL) tools (see Figure \ref{fig:dw}), a data warehouse is usually maintained seperately from the organization's operational databases \cite{sen2005comparison, 248616}. This architecture can be justified owing to the fact that operational databases are finely tuned to satisfy known OLTP requirements and functionalities which are quite different from that of on-line analytical processing (OLAP) which is supported by data warehousing. Analyzing data for decision support usually requires consolidating data from many heterogeneous sources of varying quality, or use inconsistent representations, codes and formats. Moreover, understanding trends or making predictions requires historical data, whereas operational databases store only current data. In OLAP, there's a need to support multidimensional data models and operations which requires special data organization, access methods, and implementation methods, not generally provided by DBMSs targeted for OLTP \cite{248616}. Finally, in data warehousing, query throughput and response times are more important than transaction throughput which is the major performance metric for operational databases.

\begin{figure}[!t]
\centering
\includegraphics[width=9cm]{figs/dw.png}
\caption{Data Warehousing Architecture \cite{248616}}
\label{fig:dw}
\end{figure}

@@

Choosing which indexes to create, which views to materialize, and how to partition data are hard choices. They are highly dependent on the schema used as well
as the query workload. The number of combinations of choices that can be made is very large making this task quite tedious and difficult for database
administrators. Accordingly, researchers have been inclined to find ways to automate these decisions. There were many early research efforts done in this
area(E.g.~\cite{finkelstein1988physical, hammer76selec, frank1992selec, chaudhuri1997overview, schiefer1999db2, valentin2000select}), but most of these efforts
focused on having a static workload. Accordingly, research shifted to more dynamic ways to automate physical design. Initial attempts used an dynamic, on-line
approaches(E.g.~\cite{bruno2007online, sattler2003, schnaitter2007}) where they observed the changes in the workload and adjusted the design accordingly while
attempting to predict future behavior based on the observed behavior. To avoid the problems of on-line design, researchers are also currently looking at how to
perform this dynamic physical design off-line (E.g. ~\cite{hannes2008,agrawal2006}).

A second question is how to characterize
scenarios or classes of workloads for which constrained dynamic
physical designs will be beneficial.

Similarly, IBM~\cite{} suggested a similar technique for their database with some changes such as selecting the order of
columns within an index based on the order in which the query optimizer evaluates the predicates.

Indexing in the cloud~\cite{wang2010indexing}.

Two-Tier Storage DBMS (using disk based and main memory dbs)~\cite{eo2008two}.

>>>>

\documentclass[12pt,a4paper]{article}
\usepackage{fullpage}

\usepackage{graphics}
\usepackage{color}
\usepackage[pdftex]{hyperref}
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{url}
\usepackage{multirow}
\begin{document}

@@

Relational Database Management Systems (RDBMS) have been the most popular type of Information Systems for decades. When we think of designing a database system,
we usually think of which data we need to keep track of and how will we properly organize it into meaningful tables. However, this is not the only phase in
database design. All of the designed tables need to be physically stored on disk. How these tables will be stored and accessed greatly affects the database
performance. Physical database design is concerned with the actual storage of data on disk. This means how the files are organized, and how they are accessed.
Data is stored in the form of records in files which reside on a physical disk. In this sense, we try to find the most efficient way possible to store and
access the data such that queries are answered and executed efficiently. In the traditional relational database system, this involved selecting proper indexing
techniques, partitioning the data for more efficient access, materializing the results of some queries, clustering and compressing the data, and many other
techniques \cite{fikelstein1988physical,lightstone2007physical}. The physical design aspects of relational databases have been studied in much detail. However,
over time, different types of Information Systems have emerged that have different characteristics. For example, most relational databases have been disk-based
where the primary copy of the data resides on disk. This is as opposed to Main Memory Relational Databases (MMDB) where the primary copy of the data resides in
main memory and no disk operations are needed to query the data. Such differences pose new requirements for the physical design of these systems.

@@

The rest of the survey is organized as follows. Section \ref{SEC-ELEMENTS} first explains different elements of physical database design used in traditional
disk-based databases. Section \ref{SEC-DIFFSYS} then discusses how each of the three other systems uses these elements, as applicable. Sections \ref{SEC-MMDB},
\ref{SEC-XML}, \ref{SEC-WAREHOUSES}, discuss Main Memory Databases, XML Databases, and Data Warehouses respectively. Throughout these sections, comparisons are
made as to how the different elements have been modified to fit the needs of each system. Section \ref{SEC-OPEN} mentions some of the open problems of physical
database design, and suggests possible solutions. Finally, Section \ref{SEC-CONCL} summarizes and concludes this survey.

@@

Relational Databases were first introduced by Codd \cite{codd1970relational}. Disk based Relational Database Management Systems (RDBMS) are the most common and
popular type of Information Systems. Since they have been extensively researched and have been around for many decades, we use them as the basis of our
comparison. In this section, we describe the different elements of physical design in RDBMS. These elements will also be examined in the other systems to see
how they apply to them.

@@

Btrees (short for Balanced Trees) are the most used database index structure.  The Btree is essentially similar to the traditional Binary Search Tree, but
instead of having one value per node, a B-tee can have many values per node \cite{comer1979ubiquitous}. B-trees are suitable for relational databases since the
cost of retrieval in a B-tree is at most proportional to: $log_{d}\frac{n+1}{2}$. The cost of insertion and deletion is at most proportional to $log_{d}n$ due
to the possibility of progressing back up the tree to balance it after an insertion or a deletion. Although B-trees do well in retrieval, deletion and
insertion, they do not perform well in sequential search.

Different variations of B-trees have been used. For example, B+trees are now the main method of indexing in current disk-based Relational Databases
\cite{lightstone2007physical} such as DB2, Oracle, and SQL Server. The main difference between the B+tree and the Btree is that only leaf nodes contain data
pointers in a B+tree, and leaf nodes contain pointers to each other as shown in Figure \ref{fig:b+tree}. In Prefix B+trees, instead of using the actual key
value of a record, a prefix from the key value is used to decrease the storage size needed which will in turn allow more records to be stored per node thus
decreasing the height of the tree. B+trees have been successful in relational databases since data is ultimately recorded in files. Since B+trees have a high
fanout, they allow less I/O operations to access a specific data record which is stored in a specific file.

Bitmap indices represent another type of index structures that is commonly used and cited in the literature. Using bitmaps as indexes first appeared in the
literature as part of the architecture of the Model 204 commercial database of Computer Corporation of America for the IBM 370 computer \cite{patrick1987model}.
A bitmap index structure is based on a bitmap (i.e. bitset or bit array) data structure where individual bits correspond to values in unique records. Given a
table column $C$ where a record $x$ can have one of six unique values, a traditional bitmap indexe defined for $C$ generates one bitmap vector for each of the
six different values ($B0-B5$). If the value of $C$ for record $x$ is set to 3, the corresponding bit in vector $B3$ is set to 1. Otherwise the bit is set to 0.

Bitmap indices are best used when the set of possible key values in the index is small with a large number of rows, e.g. gender attribute \cite{212001}.
Therefore, bitmap indices outperform B+trees for low cardinality attributes. The reason is that given an attribute with low cardinality, defining a B+tree index
on it means that the tree will have very few leaf nodes, each one storing a very long list of record identifiers (RIDs). On the other hand, defining a
bitmapindex on that same attribute means that such long lists or RIDs will not be stored, rather bit vectors that represent the distinct attribute values will
be stored as previously explained. This leads to a direct benefit of consuming less storage space. Moreover, queries in that case can be answered using bit-wise
logical operatios (e.g. and, or, not) which can improve query performance \cite{253268}. Another advantage of bitmap indices is the ability to perform more
efficient joins by creating bitmap join indices and using those indices to calculate the result of the join \cite{212001}. Although there are many benefits
gained from using bitmap indices, they poorly perform during updates. The reason is that when an updates is to be done to a bitmap index, the whole bit vector
for the index is locked, blocking any access to the index for other operations, then the lock is released when the update is done. That is why bitmap indices
are usually used for read-mostly databases, a famous example of which are data warehouses (Section \ref{SEC-WAREHOUSES}).

Another variation of indexing, a hash table index, often simply known as a hash index is used to organize unordered physical database tables for efficient
access \cite{lightstone2007physical}. This type of index is usually defined for keys that have unique values in the data records. Creating this index requires
passing the attribute (key) value to a hash function that maps it to a starting block address (i.e. bucket address). Afterwards, table insertions will be done
according to the hash function, as well as quering the records later.

In general, indices are a way of organizing the logical structure of a database without altering its physical structure. However, sometimes it is desirable to
store related records near each other on disk for better locality. This can be achieved through implementing a clustered index on the desired key values.
Otherwise, the created index is known as non-clustered index. The primary advantage of a clustered index is therefore the ordering of the physical data rows in
accordance with the index blocks that point to them which. Therefore, fewer data block reads are required to answer a given query. The only caveat with using a
clustered index is that there can only be one clustered index per table because the physical organization of the table must be fixed
\cite{lightstone2007physical}.

@@

In a large database system, there are usually a few complex queries that require the joining of many large tables. If these queries are frequently run, then the
database performance is likely to suffer, and users have to wait for a long time to get the results back. If we know these queries beforehand, it makes sense to
simply store the results of the queries on disk instead of recomputing them each time. Oracle 8i introduced materialized views to fit this purpose. By
precalculating the results of a complex query, and storing them in a table on disk, the new table with the results is very likely to be much smaller than the
original tables decreasing I/O access costs and thus increasing performance \cite{lightstone2007physical, chaudhuri1998overview}. Once a materialized view
exists, the user can either explicitly query the materialized view, or enter the original query, and have the query optimizer rewrite the query to use the
materialized views instead of the original tables. Many query optimization techniques have relied on rewriting queries using views (e.g.
\cite{levy1995answering, gupta1995aggregate, goldstein2001optimizing, abiteboul1998complexity}), but discussing the details of the rewriting process is beyond
the scope of this survey.

@@

Partitioning is an important aspect of physical design as it reduces table scanning time \cite{lightstone2007physical}. There are two main categories of
partitioning: horizontal partitioning and vertical partitioning \cite{agrawal2004integrating}. Horizontal partitioning divides the table into sets of table rows
where each row or record still has all the attributes of the table \cite{ceri1982horizontal, shin1985partitioning}. For example, dividing the data by date where
all data dating less than year 2000 lies in one partition while all data dating more lies in another. On the other hand, vertical partitioning reduces the width
of the dataset by storing some attributes in one table, and some in another. Usually, less frequently queried columns are partitioned together to reduce the
size of the table, especially in data warehouses where not all the columns of a table are analyzed for decision support.

The main advantage of any type of partitioning is to reduce the amount of time it takes to scan a table which in turn improves performance
\cite{lightstone2007physical}. For example, if a table has 1000 records, and a query is only using the first 20 records then having these 20 records in a
partition containing 50 records will save time as opposed to examining all the 1000 records. Horizontal partitioning is more widely used, and is implemented in
all current DBMSes \cite{mysqlpartitioning, sqlsrvpartitioning, sybasepartitioning, postgrepartitioning, oraclepartitioning, db2partitioning} which explains
why there exists many flavors of horizontal paritioning techniques. There are three major types of horizontal partitioning: range partitioning
\cite{dewitt1986gamma}, list partitioning \cite{lightstone2007physical} (which is a variation of range partitioning), and hash partitioning
\cite{lightstone2007physical}. Of course, one can mix and match those partitioning schemes and come up with a composite partitioning that fits a particular
situation \cite{lightstone2007physical}. Table \ref{tab:horpartition} shows different categories of horizontal partitioning, and the situations they are suited
to. Combining partitioning and indexing can often lead to better performance. For example, if List partitioning is used to divide a table by regions and then a
bitmap index is used on that table, finding records in the same region or state can be much faster. Finally, partitioning the data into different physical
devices to achieve parallelism when accessing it can greatly increase the performance of the database.

\begin{table}[!t]
\scalebox{0.7}{
 \begin{tabular}{p{2cm}||p{4cm}|p{4cm}|p{4cm}|p{4cm}}
& \textbf{Range Partitioning} & \textbf{List Partitioning} & \textbf{Hash Partitioning} & \textbf{Composite Partitioning}\\
\hline\hline
\textbf{How?} & If key falls in range & List of keys & Hash function& Hybrid\\
\hline
\textbf{Why?} & Disk limits & Group point values & Range/List are N/A & According to needs\\
\hline
\textbf{Benefits} & \multicolumn{4}{c}{Partition elimination, Improved administration, Fast roll-in/out, and Implicit clustering }\\
\hline
\textbf{Example} & \begin{itemize}
                    \item Zip code 
		    \item Time range
                   \end{itemize} &
		    \begin{itemize}
		    \item Regions
		    \item States
		    \end{itemize} &
		    \begin{itemize}
		    \item EmployeeId
		    \item ProductId
		    \end{itemize} &
		    Hash then List \\
\hline
\textbf{RDBMS} & MySQL, DB2, Oracle, PostgreSQL, SQL Server & MySQL, Oracle, PostgreSQL, SQL Server & MySQL, Oracle & MySQL, Oracle\\
\end{tabular}
}
\caption{Types of horizontal partitioning}
\label{tab:horpartition}
\end{table}

@@

A Main Memory Database (MMDB) (sometimes referred to as in-memory database) is one where the primary copy of the data resides in the main memory of the system
rather than on a
disk~\cite{garcia1992main}. It should be noted that this is different from caching. Disk based database systems use the main memory to cache query results.
However, MMDB's primary copy of the data resides in main memory. Accessing data in main memory is faster than disk, and so MMDBs usually have better
performance that disk based systems. However, main memory is more expensive and the size is limited in comparison to disk. However, MMDBs are essential to use
where some of the collected data cannot, or will not, be stored on disk in the first place. For example, phone switch data may be needed for reasoning with
other stored data, but it is captured in real time and will not necessarily be stored on disk. MMDB have many implementation challenges that have been addressed
throughout the research community. These challenges will be discussed throughout this section. %Table~\ref{tab:mmdbsumm} summarizes
the physical design elements of MMDBs.

\begin{table}[t!]
\centering
\scalebox{0.9}{
\begin{tabular}{p{3cm}|p{6cm}|p{6cm}}
\hline
 \textbf{Physical Design Element} & \textbf{Desired Features} & \textbf{Methods/Structures Used}\\
\hline\hline
Index Structures & \begin{itemize}
                    \item No need to store actual values in index \item Larger node size that is aligned with cache line size for cache consciousness
                   \end{itemize} & \begin{itemize}
                                    \item B Tree
\item B+ Tree
\item CSB+ Tree
\item pB+ Tree
\item T Tree
\item CST Tree
\item CSS Tree
                                   \end{itemize}\\
\hline
Materialized Views & \begin{itemize} \item Not highly utilized since processing and memory access is cheap \end{itemize}& N/A\\
\hline
Partitioning & \begin{itemize} \item Only needed in secondary storage to speed up reloading in case of a crash \end{itemize} & \begin{itemize}
                                                                                            \item Horizontal Partitioning
\item Vertical Partitioning
                                                                                           \end{itemize}\\
\hline
Clustering & \begin{itemize} \item Not needed since random or dispersed main memory access is not more expensive than sequential access \end{itemize}
& N/A

@@

In this section, we talk about three families of index structures used in MMDB: B Trees, T Trees, and Binary Search Trees. There are more recent index
structures developed for MMDB such as J+ Trees~\cite{luan2009prefetching}, but which have not gained much popularity, and so we do not discuss them in our
survey. %Table~\ref{tab:mmdbindexsumm} summarizes the features of the different index structures discussed in this section with respect to MMDB. The T Tree is
the most used index for MMDB (E.g. MySQL CLuster). However, there are other indexes used as well. For example, IBM's solidDB uses Trie (or prefix tree) for its
indexing. For their purposes, it served as a good index for main memory since it eliminates the need for many key comparisons~\cite{ibmsoliddb}.

 \begin{table}[t!]
\centering
\scalebox{0.6}{
\begin{tabular}{p{2cm}|p{6cm}|p{6cm}}
\hline
 \textbf{Index Structure} & \textbf{Features} & \textbf{Cache Consciousness}\\
\hline\hline
B Trees &
\begin{itemize}\item Good storage utilization
\item Reasonably quick searching
\item Fast updating
\end{itemize} 
& \begin{itemize}\item Reasonable cache behavior if node fits in cache line\end{itemize}\\
\hline
B+ Trees & 
\begin{itemize}
\item Reasonably quick searching
\item Fast updating
\end{itemize} &
\begin{itemize}
\item Reasonable cache behavior if node fits in cache line\end{itemize}\\
\hline
CSB+ Trees & \begin{itemize}
              \item Same features of B+ Trees
\item Stores child nodes in an array. Stores pointer to first child only and uses index calculation to get the rest
             \end{itemize} &
\begin{itemize}
 \item Good cache performance due to improved locality \& lower tree height (because of larger node size)
\end{itemize}\\
\hline

@@

\subsection{Data Warehouses}
\label{SEC-WAREHOUSES}
A data warehouse is a subject-oriented, integrated, time-variant, and non-volatile collection of data and decision support technologies, aimed at enabling the
knowledge worker (executive, manager, analyst) to make better and faster decisions \cite{560407, chaudhuri1997overview}. More than half of IT executives named
Data warehousing as the highest-priority post-millennium project for them \cite{sen2005comparison}. The value of data warehousing for an organization depends on
the organization's need for reliable, consolidated, unique and integrated reporting and analysis of its data, at different levels of aggregation.

Data warehousing has been shown useful in many industries: manufacturing (for order shipment and customer support), retail (for user profiling and inventory
management), financial services (for claims analysis, risk analysis, credit card analysis, and fraud detection), transportation (for fleet management),
telecommunications (for call analysis and fraud detection), utilities (for power usage analysis), and healthcare (for outcomes analysis)
\cite{chaudhuri1997overview}.

@@

Although data in a data warehouse is extracted and loaded from multiple on-line transaction processing (OLTP) data sources (including DB2, Oracle, IMS
databases, and flat files) using Extract, Transfer, and Load (ETL) tools (see Figure \ref{fig:dw}), a data warehouse is usually maintained seperately from the
organization's operational databases \cite{sen2005comparison, chaudhuri1997overview}. This architecture can be justified owing to the fact that operational
databases are finely tuned to satisfy known OLTP requirements and functionalities which are quite different from that of on-line analytical processing (OLAP)
which is supported by data warehousing. Analyzing data for decision support usually requires consolidating data from many heterogeneous sources of varying
quality, or use inconsistent representations, codes and formats. Moreover, understanding trends or making predictions requires historical data, whereas
operational databases store only current data. In OLAP, there's a need to support multidimensional data models and operations which requires special data
organization, access methods, and implementation methods, not generally provided by DBMSs targeted for OLTP \cite{chaudhuri1997overview}. Finally, in data
warehousing, query throughput and response times are more important than transaction throughput which is the major performance metric for operational databases.

\begin{figure}[!t]
\centering
\includegraphics[width=9cm]{figs/dw.png}
\caption{Data Warehousing Architecture \cite{chaudhuri1997overview}}
\label{fig:dw}
\end{figure}

@@

Automating the physical database design choices has been a very attractive area of research that is still receiving lots of attention until now. Choosing which
indexes to create, which views to materialize, and how to partition data are hard choices. They are highly dependent on the schema used as well
as the query workload. The number of combinations of choices that can be made is very large making this task quite tedious and difficult for database
administrators. Accordingly, researchers have been inclined to find ways to automate these decisions. There were many early research efforts done in this
area(E.g.~\cite{finkelstein1988physical, hammer76selec, frank1992selec, chaudhuri1997overview, schiefer1999db2, valentin2000select}), but most of these efforts
focused on having a static workload. Accordingly, research shifted to more dynamic ways to automate physical design. Initial attempts used an dynamic, on-line
approaches(E.g.~\cite{bruno2007online, sattler2003, schnaitter2007}) where they observed the changes in the workload and adjusted the design accordingly while
attempting to predict future behavior based on the observed behavior. To avoid the problems of on-line design, researchers are also currently looking at how to
perform this dynamic physical design off-line (E.g. ~\cite{hannes2008,agrawal2006}). Additionally, finding better evaluation techniques to compare different
automatic physical design decision is currently being studied~\cite{gebaly2008}.

Other problems which are currently being researched include improving I/O operations in database systems by using Solid State Drives (SSD) instead of of Hard
Disk Drives (HDD)~\cite{lee2009, du2009, lee2008}. SSDs have faster read performance, but perform poorly in random writes. However, for certain applications in
databases and for databases where random writes are not as frequent, SSDs might be suitable. Using SDDs also seems promising when used for
transaction logs, and intermediate results of sorting and joining~\cite{lee2008}.

Just like all other research areas are moving to the cloud, database systems are moving there too. Accordingly, lots of current research focuses on adapting
storage techniques and index structures to the cloud. For example, Wang et al. propose a new index structure for indexing mult-dimensional data in the
cloud~\cite{wang2010indexing}. Lots of work is currently being done in this area. For example, Wu et al.~\cite{wu2010efficient} propose a modified B+tree index
for datasets stored in each compute node. The exact requirements and solutions for data management in the cloud are still being
researched~\cite{abadi2009data}, and so this is an area that is expected to receive more attention in the future as cloud computing becomes more prevalent.

This is by no means an exhaustive list of the problems still being researched in physical database design, but they provide a flavor of the challenges still
being faced. As shown, as new technologies such as SSDs or cloud computing environments emerge, database systems will need to adapt themselves to either make
use of them or to be able to cope with the new requirements they pose.
